
title: "Diabetes Prediction: Logistic Regression Analysis"
author: "Kaushal"
date: "24-03-2026"
output:html_document
  

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```



# Introduction

This report presents a logistic regression analysis of the **Diabetes Prediction Dataset** sourced from [Kaggle](https://www.kaggle.com/datasets/iammustafatz/diabetes-prediction-dataset).

**Study Design:** Cross-sectional observational study  
**Study Population:** ~100,000 patients with demographic, lifestyle, and clinical information  
**Outcome:** Binary diabetes status (0 = No Diabetes, 1 = Diabetes)  

The aim is to build and evaluate a logistic regression model to predict the probability of diabetes based on 8 clinical and demographic predictors.



# Setup and Data Loading

```{r libraries}
library(tidyverse)
library(caret)
library(pROC)
library(ggplot2)
library(gridExtra)
library(knitr)
library(kableExtra)
library(ResourceSelection) 
```

```{r load-data}

df <- read.csv("diabetes_prediction_dataset.csv", stringsAsFactors = FALSE)


dim(df)
head(df)
```



# Data Inspection and Preprocessing

## Variable Overview

```{r variable-overview}
str(df)
summary(df)
```

## Missing Data Assessment

```{r missing-data}
missing_summary <- data.frame(
  Variable = names(df),
  Missing_Count = colSums(is.na(df)),
  Missing_Percent = round(colSums(is.na(df)) / nrow(df) * 100, 2)
)

kable(missing_summary, row.names = FALSE,
      caption = "Missing Data Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Encoding and Cleaning

```{r preprocessing}
df <- df %>%
  mutate(
    gender = case_when(
      gender == "Male"   ~ 1,
      gender == "Female" ~ 0,
      TRUE               ~ NA_real_
    ),
    smoking_history = case_when(
      smoking_history == "never"   ~ 0,
      smoking_history == "former"  ~ 1,
      smoking_history == "current" ~ 2,
      smoking_history == "ever"    ~ 1,
      smoking_history == "not current" ~ 1,
      TRUE                         ~ NA_real_
    )
  )


df_clean <- df %>% drop_na()

cat("Observations after cleaning:", nrow(df_clean), "\n")
cat("Diabetes cases:", sum(df_clean$diabetes), "\n")
cat("Prevalence:", round(mean(df_clean$diabetes) * 100, 2), "%\n")
```



# Exploratory Data Analysis

## Outcome Distribution

```{r outcome-dist}
ggplot(df_clean, aes(x = factor(diabetes), fill = factor(diabetes))) +
  geom_bar(width = 0.5, show.legend = FALSE) +
  scale_fill_manual(values = c("blue", "red")) +
  scale_x_discrete(labels = c("No Diabetes", "Diabetes")) +
  labs(title = "Distribution of Diabetes Outcome",
       x = "Diabetes Status", y = "Count") +
  theme_minimal(base_size = 13)
```




# Model Building

## Train-Test Split

```{r train-test-split}
set.seed(42)

train_index <- createDataPartition(df_clean$diabetes, p = 0.8, list = FALSE)
train_data  <- df_clean[train_index, ]
test_data   <- df_clean[-train_index, ]

cat("Training set:", nrow(train_data), "observations\n")
cat("Test set:", nrow(test_data), "observations\n")
```

## Logistic Regression Model

```{r logistic-model}
model <- glm(
  diabetes ~ age + gender + hypertension + heart_disease +
             smoking_history + bmi + HbA1c_level + blood_glucose_level,
  data   = train_data,
  family = binomial(link = "logit")
)

summary(model)
```

## Odds Ratios and Confidence Intervals

```{r odds-ratios}
or_table <- data.frame(
  Predictor = names(coef(model)),
  Estimate  = round(coef(model), 4),
  OR        = round(exp(coef(model)), 4),
  CI_Lower  = round(exp(confint(model))[, 1], 4),
  CI_Upper  = round(exp(confint(model))[, 2], 4),
  P_value   = round(summary(model)$coefficients[, 4], 4)
)

kable(or_table, row.names = FALSE,
      caption = "Logistic Regression: Odds Ratios and 95% Confidence Intervals") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(or_table$P_value < 0.05), bold = TRUE, color = "white",
           background = "green")
```



# Model Evaluation

## Predictions on Test Set

```{r predictions}
test_probs <- predict(model, newdata = test_data, type = "response")
test_preds <- ifelse(test_probs > 0.5, 1, 0)
```

## Confusion Matrix

```{r confusion-matrix}
cm <- confusionMatrix(factor(test_preds), factor(test_data$diabetes),
                      positive = "1")
print(cm)
```

## ROC Curve and AUC

```{r roc-curve}
roc_obj <- roc(test_data$diabetes, test_probs)

plot(roc_obj,
     col = "grey", lwd = 2.5,
     main = paste("ROC Curve | AUC =", round(auc(roc_obj), 3)),
     print.auc = TRUE)
abline(a = 0, b = 1, lty = 2, col = "grey60")
```

## Calibration: Hosmer-Lemeshow Test

```{r hosmer-lemeshow}
hl_test <- hoslem.test(test_data$diabetes, test_probs, g = 10)
print(hl_test)
```

A non-significant p-value (p > 0.05) suggests good model calibration.

## Predicted Probability Distribution

```{r pred-prob-dist}
pred_df <- data.frame(
  predicted_prob = test_probs,
  actual         = factor(test_data$diabetes,
                          labels = c("No Diabetes", "Diabetes"))
)

ggplot(pred_df, aes(x = predicted_prob, fill = actual)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("blue", "orange"), name = "Actual Status") +
  labs(title = "Distribution of Predicted Probabilities",
       x = "Predicted Probability of Diabetes",
       y = "Count") +
  theme_minimal(base_size = 12)
```


# Model Performance Summary

```{r performance-summary}
perf_table <- data.frame(
  Metric    = c("Accuracy", "Sensitivity (Recall)", "Specificity",
                "Positive Predictive Value", "Negative Predictive Value",
                "AUC-ROC"),
  Value     = c(
    round(cm$overall["Accuracy"], 4),
    round(cm$byClass["Sensitivity"], 4),
    round(cm$byClass["Specificity"], 4),
    round(cm$byClass["Pos Pred Value"], 4),
    round(cm$byClass["Neg Pred Value"], 4),
    round(auc(roc_obj), 4)
  )
)

kable(perf_table, row.names = FALSE,
      caption = "Logistic Regression Model Performance Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```


# Discussion
No surprises — HbA1c and blood glucose were the strongest predictors, exactly what you'd expect clinically. BMI and age also contributed. Hypertension and heart disease were present but likely overlap with other variables.
The calibration test flagged a slight issue (p = 0.018), so the model's predicted probabilities aren't perfectly reliable.
Also worth noting — it's cross-sectional data, so we can't claim causation, just association.
Overall, solid baseline model. Could be improved by exploring interactions or trying more advanced algorithms.

# Conclusion
1.The logistic regression model did a great job predicting diabetes with 94.85% accuracy and an AUC of 0.954 — which is honestly impressive.
2.The two biggest factors driving predictions were HbA1c level and blood glucose level, which makes complete sense medically. Age and BMI also played a role.
3.The only real weak spot is sensitivity at 63% — meaning it misses about 1 in 3 actual diabetes cases, which isn't ideal if you're using this in a medical setting.
4.It's a solid, well-performing model. Good enough for a baseline, but you'd want to improve how it catches positive cases before trusting it in the real world.
