---
title: "Predictive modelling assignment1"
author: "Kaushal Mayur"
date: "2026-03-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```

---

# Introduction

This report presents a logistic regression analysis of the **Diabetes Prediction Dataset** sourced from [Kaggle](https://www.kaggle.com/datasets/iammustafatz/diabetes-prediction-dataset).

**Study Design:** Cross-sectional observational study  
**Study Population:** ~100,000 patients with demographic, lifestyle, and clinical information  
**Outcome:** Binary diabetes status (0 = No Diabetes, 1 = Diabetes)  

The aim is to build and evaluate a logistic regression model to predict the probability of diabetes based on 8 clinical and demographic predictors.

---

# Setup and Data Loading

```{r libraries}
library(tidyverse)
library(caret)
library(pROC)
library(ggplot2)
library(gridExtra)
library(knitr)
library(kableExtra)
library(ResourceSelection) 
```

```{r load-data}

df <- read.csv("diabetes_prediction_dataset.csv", stringsAsFactors = FALSE)


dim(df)
head(df)
```

---

# Data Inspection and Preprocessing

## Variable Overview

```{r variable-overview}
str(df)
summary(df)
```

## Missing Data Assessment

```{r missing-data}
missing_summary <- data.frame(
  Variable = names(df),
  Missing_Count = colSums(is.na(df)),
  Missing_Percent = round(colSums(is.na(df)) / nrow(df) * 100, 2)
)

kable(missing_summary, row.names = FALSE,
      caption = "Missing Data Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Encoding and Cleaning

```{r preprocessing}
# Recode gender: keep Male/Female, set 'Other' to NA
df <- df %>%
  mutate(
    gender = case_when(
      gender == "Male"   ~ 1,
      gender == "Female" ~ 0,
      TRUE               ~ NA_real_
    ),
    # Recode smoking history to ordinal numeric
    smoking_history = case_when(
      smoking_history == "never"   ~ 0,
      smoking_history == "former"  ~ 1,
      smoking_history == "current" ~ 2,
      smoking_history == "ever"    ~ 1,
      smoking_history == "not current" ~ 1,
      TRUE                         ~ NA_real_
    )
  )

# Remove rows with any remaining NAs
df_clean <- df %>% drop_na()

cat("Observations after cleaning:", nrow(df_clean), "\n")
cat("Diabetes cases:", sum(df_clean$diabetes), "\n")
cat("Prevalence:", round(mean(df_clean$diabetes) * 100, 2), "%\n")
```

---

# Exploratory Data Analysis

## Outcome Distribution

```{r outcome-dist}
ggplot(df_clean, aes(x = factor(diabetes), fill = factor(diabetes))) +
  geom_bar(width = 0.5, show.legend = FALSE) +
  scale_fill_manual(values = c("#2196F3", "#F44336")) +
  scale_x_discrete(labels = c("No Diabetes", "Diabetes")) +
  labs(title = "Distribution of Diabetes Outcome",
       x = "Diabetes Status", y = "Count") +
  theme_minimal(base_size = 13)
```

## Continuous Predictor Distributions by Outcome

```{r continuous-dist}
cont_vars <- c("age", "bmi", "HbA1c_level", "blood_glucose_level")

plots <- lapply(cont_vars, function(var) {
  ggplot(df_clean, aes_string(x = var, fill = "factor(diabetes)")) +
    geom_density(alpha = 0.6) +
    scale_fill_manual(values = c("#2196F3", "#F44336"),
                      labels = c("No Diabetes", "Diabetes"),
                      name = "") +
    labs(title = var, x = var, y = "Density") +
    theme_minimal(base_size = 11)
})

grid.arrange(grobs = plots, ncol = 2)
```

## Binary Predictors: Diabetes Prevalence

```{r binary-predictors}
binary_vars <- c("hypertension", "heart_disease")

prev_table <- df_clean %>%
  select(all_of(binary_vars), diabetes) %>%
  pivot_longer(-diabetes, names_to = "variable", values_to = "value") %>%
  group_by(variable, value) %>%
  summarise(prevalence = round(mean(diabetes) * 100, 2), .groups = "drop")

ggplot(prev_table, aes(x = factor(value), y = prevalence, fill = variable)) +
  geom_col(position = "dodge", width = 0.5) +
  facet_wrap(~variable, scales = "free_x") +
  scale_fill_manual(values = c("#4CAF50", "#FF9800")) +
  labs(title = "Diabetes Prevalence by Binary Predictors",
       x = "Value (0 = No, 1 = Yes)", y = "Prevalence (%)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## Correlation Heatmap (Continuous Variables)

```{r correlation}
num_vars <- df_clean %>%
  select(age, bmi, HbA1c_level, blood_glucose_level, diabetes)

cor_matrix <- round(cor(num_vars, use = "complete.obs"), 2)

# Melt and plot
cor_df <- as.data.frame(as.table(cor_matrix))
names(cor_df) <- c("Var1", "Var2", "Correlation")

ggplot(cor_df, aes(Var1, Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Correlation), size = 3.5) +
  scale_fill_gradient2(low = "#2196F3", mid = "white", high = "#F44336",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Correlation Heatmap", x = "", y = "") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```